---
title: "Week 3 exercises"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading common packages

```{r}

```

## Exercise 1: summary statistics on the penguins dataset

```{r}
data(penguins, package="palmerpenguins")
```

Work with the penguins dataset.

1. Obtain the summary statistics on bill length, flipper length, and body mass. Also count the occurrences of the two sexes.

2. For each unique combination of species and residing island of penguins, find the mean and standard deviation in body mass.

## Exercise 2: hypothesis testing

1. Suppose you reject the null hypothesis in a two-tailed test. Will you reject the corresponding null hypothesis in a one-tailed test, at the same significance level 𝛼?

2. You are trying to determine if there’s real difference in the mean between two populations A and B. Suppose the difference is real, by increasing the sample size, would you expect…

  a. Higher statistical power?
  b. Higher effect size?
  c. Lower p-value?

## Exercise 3: Salish Sea salinity (simulated data)

Samples of seawater at near-surface depth ($d$ < 5 m) were collected at 3 different locations in the Salish Sea. The data can be found in the file `salish_cruise_simulated_1.csv`. Perform an appropriate t-test on the difference in mean salinity between the Tacoma Narrows samples and the Main Basin samples

## Exercise 4: Salish Sea salinity revisited

Going back to the Salish Sea water samples in Exercise #3. Determine if t-test was appropriate.

## Exercise 5: More Salish Sea data (simulated data)

More Salish Sea data seawater observations were taken at the 3 locations as in example 1, and the results are collected in the file salish_cruise_simulated_2.csv. This time, each set of data was collected over a broader range of location.

1. Check if t-test is still appropriate for comparisons between the Tacoma Narrows and the Main Basin samples

2. Perform the Wilcoxon-Mann-Whitney rank sum test on the comparison of salinity between the Main Basin and the Tacoma Narrows. Is the difference statistically significant (=discernible)?

3. In addition, compute the effect size of the observed difference

## (Optional) Feedback questions

1. Are there any topic(s) from this week that still confuse you?

2. Which activities work and which doesn't? What do you want to see change and what do you want to keep?

