---
title: "Summary statistics in R"
author: "Wing-Ho Ko"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the packages we'll need

```{r}
library(tidyverse)
```

## Summary statistics

Keep a simpler dataframe for the summary statistics examples below

```{r}
data(storms, package="dplyr")
```

Simplify the dataset a little bit

```{r}
storms_sim <- storms |>
  select(name:hour | status:pressure)
```

```{r}
glimpse(storms_sim)
```

Calculate basic summary statistics using `summary()`

```{r}
summary(storms_sim)
```

Calculate the average category directly. Notice that in this case NA propagates....

```{r}
mean(storms$category)
```

To omit NA's, add the named argument `na.rm=TRUE`

```{r}
category_mean <- mean(storms$category, na.rm=TRUE)
print(category_mean)
```

Similarly, to obtain standard deviation we may do (note that R computes the standard deviation that derives from the unbiased point estimator for variance):

```{r}
category_sd <- sd(storms$category, na.rm=TRUE)
print(category_sd)
```

For median, use `median()` instead of `mean()`

```{r}
median(storms$category, na.rm=TRUE)
```

More generally, use `quantile()` for quantiles:

```{r}
quantile(storms$category, 0.25, na.rm=TRUE)
```

We can bundle multiple calculations of summary statistics using `summarize()`

```{r}
storms_sim |> summarise(
  mean_wind = mean(wind, na.rm=TRUE),
  sd_wind = sd(wind, na.rm=TRUE),
  mean_pres = mean(pressure, na.rm=TRUE),
  sd_pres = sd(pressure, na.rm=TRUE)
)
```

## Grouped summary statistics

Calculate the monthly average for wind speed

```{r}
grouped <- storms |> group_by(month) |>
  summarize(
    wind_mean = mean(wind, na.rm=TRUE),
    wind_sd = sd(wind, na.rm=TRUE),
  )

grouped
```

Calculate the for wind speed per year-month

```{r}
grouped2 <- storms |> group_by(year, month) |>
  summarize(
    wind_mean = mean(wind, na.rm=TRUE),
    wind_sd = sd(wind, na.rm=TRUE),
  ) |> ungroup()

grouped2
```
