---
title: "t-test demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(broom)
library(MKpower)
library(effectsize)
```

## Orca example (simulated data)

Suppose we have tagged 20 orcas in the Washington shore, and we recorded their weights one year apart. The data is stored in the orcas_simulated.csv file in the data folder

Perform an appropriate t-test to check if there’s discernable difference between the weights of orcas between the two measurement

**t-test to perform**: the observations are paired. Thus a paired t-test is needed

Load the data and inspect it

```{r}
orcas <- read_csv("data/orca_simulated.csv")
glimpse(orcas)
```

Subset data that can be used as input to t.test()

```{r}
orcas_1 <- orcas |> filter(year==2023) |> arrange(tag)
orcas_2 <- orcas |> filter(year==2024) |> arrange(tag)

orcas_wt_1 <- orcas_1$weight_kg
orcas_wt_2 <- orcas_2$weight_kg
```

Perform t.test. Defaulted to printing out the results

```{r}
t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided", paired=TRUE)
```

Assign the `t.test()` results to variable

```{r}
orcas_t <- t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided", paired=TRUE)
```

Extract the p-value programatically:

```{r}
print(orcas_t$p.value)
```

Extract the point estimate and confidence interval programatically:

```{r}
print(orcas_t$estimate)
```

```{r}
print(orcas_t$conf.int)
```

Alternatively, use `tidy()` from the `broom` package to create a tidy data frame from the t test output

```{r}
tidy(orcas_t)
```


## t-test revisited

Run different t-tests against the orca data

```{r}
orcas_t1 <- t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided", paired=TRUE)
orcas_t2 <- t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided")
```

Print the different p-values obtained. Observed that p-values become worse with paired vs (unpaired pooled) vs (unpaired unpooled)

```{r}
print(c(orcas_t1$p.value, orcas_t2$p.value))
```

### Statistical power

Check the statistical power assuming that the difference is "real"

```{r}
orcas_diff <- mean(orcas_wt_1) - mean(orcas_wt_2)
orcas_diff_sd <- sd(orcas_wt_1 - orcas_wt_2)

power.t.test(n=20, delta=orcas_diff, sd=orcas_diff_sd, sig.level=0.05, type="paired")
```

If instead we use an unpaired t-test (there `power.welch.t.test()` function comes from the package `MKpower`):

```{r}
orcas_diff <- mean(orcas_wt_1) - mean(orcas_wt_2)
orcas_sd_1 <- sd(orcas_wt_1)
orcas_sd_2 <- sd(orcas_wt_2)

power.welch.t.test(n=20, delta=orcas_diff, sd1=orcas_sd_1, sd2=orcas_sd_2)
```

Asking the question differently: how big does `n` needs to be for power = 0.80?

```{r}
power.t.test(
  power=0.80, delta=orcas_diff, sd=orcas_diff_sd, sig.level=0.05, 
  type="paired"
)
```

### Q-Q plot, skewness and kurtosis

Use the precipitation data from SeaTac airport, year 2000 and year 2017. Focus only on wet-day data (i.e., days where PRCP > 0)

```{r}
seatac2000 <- read_csv("data/SeaTac_prcp_2000.csv")
seatac2017 <- read_csv("data/SeaTac_prcp_2017.csv")

seatac2000 <- seatac2000 |> filter(PRCP > 0) |> select(DATE | PRCP)
seatac2017 <- seatac2017 |> filter(PRCP > 0) |> select(DATE | PRCP)
```

```{r}
length(seatac2000$PRCP)
```

```{r}
length(seatac2017$PRCP)
```

While our sample size is large. Weather is correlated events, so perhaps we should be more conservative and treat the data as moderately sized.

Check if the normality assumption is acceptable for the two sets of data:

```{r}
ggplot(seatac2000, aes(sample=PRCP)) + geom_qq() + geom_qq_line()
```

```{r}
ggplot(seatac2017, aes(sample=PRCP)) + geom_qq() + geom_qq_line()
```

Both data have heavy right skew, as can be checked from computing skewness:

```{r}
e1071::skewness(seatac2000$PRCP, type=2)
```

```{r}
e1071::skewness(seatac2017$PRCP, type=2)
```

We can also check the kurtosis:

```{r}
e1071::kurtosis(seatac2000$PRCP, type=2)
```

```{r}
e1071::kurtosis(seatac2017$PRCP, type=2)
```

So we should probably refrain from using t-test


### Wilcoxon test

Let's perform a Wilcoxon test instead:

```{r}
wilcox.test(seatac2000$PRCP, seatac2017$PRCP)
```

We see that the Wilcoxon test is somewhat inconclusive. There are apparent differences, but not highly statistically significant.

### Effect size

Let's also consider effect size:

```{r}
cohens_d(seatac2000$PRCP, seatac2017$PRCP, pooled=FALSE)
```

We see that Cohen's d that is between 0.2 and 0.5, so it the difference is small in realistic terms.
